The Efficient Handling of Guards in the Design of RPython's Tracing JIT

### 2012-08-15 talk/vmil2012/paper.tex
More small improvements

\begin{abstract}
Tracing just-in-time (JIT) compilers record linear control flow paths,
inserting operations called guards at points of possible divergence. These
operations occur frequently generated traces and therefore it is important to
design and implement them carefully to find the right trade-off between
execution speed, deoptimization,
and memory overhead.  In this paper we describe the design decisions about
guards taken in the implementation of the RPython tracing JIT. Furthermore we
measure various properties of guards.
% \o/
\end{abstract}

%___________________________________________________________________________
\section{Introduction}

@@

There are several aspects to consider
in the design and optimization of guards, the first aspect is that due to the
large number of guards the memory overhead related to storing the information
needed for deoptimization should be kept low. A second aspect is that
successfully checking guards, i.e. not leaving the compiled trace, – which is
the common case – should be a cheap operation to execute favouring the on-trace
execution speed in contrast to the deoptimization case where the state has to
be rebuilt using the stored information. These constraints and trade-offs are
what make the design and optimization of guards an important and non-trivial
aspect of the low-level design of a tracing just-in-time compiler.

%Section~\ref{sec:Evaluation} presents Figures about the absolute number of
%operations for each benchmark, and the overhead produced by the information
%stored at the different levels for the guards
In this paper we want to substantiate the aforementioned observations and
describe based on them the reasoning behind the implementation of guards in
RPython's tracing just-in-time compiler. The contributions of this paper are:
\begin{itemize}
  \item An analysis and benchmark of guards in the context of RPython's tracing JIT,
  %An analysis of guards in the context of RPython's tracing JIT to
  %substantiate the aforementioned observation, based on a set of benchmarks,
  \item detailed measurements about the frequency and the
  overhead associated with guards, and
  \item a description about how guards are implemented in the high\-
  and low-level components of the JIT and a description of the rationale behind the design.
\end{itemize}

@@

After profiling identifies an interesting
path, tracing is started thus recording all operations that are executed on this
path. This includes inlining functional calls.
As in most compilers, tracing JITs use an intermediate representation to
store the recorded operations, which is typically in SSA
form~\cite{cytron_efficiently_1991}. Since tracing follows actual execution the
code that is recorded
represents only one possible path through the control flow graph. Points of
divergence from the recorded path are marked with special operations called
\emph{guards}, these operations ensure that assumptions valid during the
tracing phase are still valid when the code has been compiled and is executed.
In the case of dynamic languages, guards are also used to encode type checks
that come from optimistic type specialization by recording the types of
variables seen during tracing.
After a trace has been recorded it is optimized and then compiled to platform
specific machine code.

When the check of a guard fails, the execution of the machine code must be
stopped and the control is returned to the interpreter, after the interpreter's
state has been restored. If a particular guard fails often a new trace is
recorded starting from the guard. We will refer to this kind of trace as a
\emph{bridge}. Once a bridge has been traced it is attached to the
corresponding guard by patching the machine code. The next time the guard fails
the bridge will be executed instead of leaving the machine code.

@@

To do this reconstruction it is necessary to take the values of the SSA
variables of the trace and build interpreter stack frames.  Tracing
aggressively inlines functions, therefore the reconstructed state of the
interpreter can consist of several interpreter frames.

If a guard fails often enough, a trace is started from it
forming a trace tree.
When that happens another use case of resume data
is to construct the tracer state.
After the bridge has been recorded and compiled it is attached to the guard.
If the guard fails later the bridge is executed. Therefore the resume data of
that guard is no longer needed.

There are several forces guiding the design of resume data handling.
Guards are a very common operations in the traces.
However, a large percentage of all operations
are optimized away before code generation.
Since there are a lot of guards
the resume data needs to be stored in a very compact way.
On the other hand, tracing should be as fast as possible,
so the construction of resume data must not take too much time.

@@

Every time a guard is recorded during tracing
the tracer attaches preliminary resume data to it.
The data is preliminary in that it is not particularly compact yet.
The preliminary resume data takes the form of a stack of symbolic frames.
The stack contains only those interpreter frames seen by the tracer.
The frames are symbolic in that the local variables in the frames
do not contain values.
Instead, every local variables contains the SSA variable of the trace
where the value would later come from, or a constant.

@@

After tracing has been finished the trace is optimized.
During optimization a large percentage of operations can be removed.
In the process the resume data is transformed into its final, compressed form.
The rationale for not compressing the resume data during tracing
is that a lot of guards will be optimized away.
For them, the compression effort would be lost.

@@

Guards interact with optimizations in various ways.
Using many classical compiler optimizations the JIT tries to remove as many
operations, and therefore guards, as possible.
In particular guards can be removed by subexpression elimination.
If the same guard is encountered a second time in the trace,
the second one can be removed.
This also works if a later guard is weaker
and hence implied by an earlier guard.

@@

Adding this additional information is done as follows:
So far, every variable in the symbolic frames
contains a constant or an SSA variable.
After allocation removal the variables in the symbolic frames can also contain
``virtual'' objects.
These are objects that were not allocated so far,
because the optimizer removed their allocation.
The structure of the heap objects that have to be allocated on guard failure
is described by the virtual objects stored in the symbolic frames.
To this end, the content of every field of the virtual object is described
in the same way that the local variables of symbolic frames are described.
The fields of the virtual objects can therefore be SSA variables, constants
or other virtual objects.
They are encoded using the same compact two-byte representation
as local variables.

During the storing of resume data virtual objects are also shared
between subsequent guards as much as possible.
The same observation as about frames applies:
Quite often a virtual object does not change from one guard to the next.
Then the data structure is shared.

@@

After optimization the resulting trace is handed over to the platform specific
backend to be compiled to machine code. The compilation phase consists of two
passes over the lists of instructions, a backwards pass to calculate live
ranges of IR-level variables and a forward pass to emit the instructions. During
the forward pass IR-level variables are assigned to registers and stack
locations by the register allocator according to the requirements of the
emitted instructions.  Eviction/spilling is performed based on the live range
information collected in the first pass. Each IR instruction is transformed
into one or more machine level instructions that implement the required
semantics. Operations without side effects whose result is not used are not
emitted. Guards instructions are transformed into fast checks at the machine
code level that verify the corresponding condition.  In cases the value being
checked by the guard is not used anywhere else the guard and the operation
producing the value can often be merged, further reducing the overhead of the guard.
Figure \ref{fig:trace-compiled} shows how the \texttt{int\_eq} operation
followed by a \texttt{guard\_false} from the trace in Figure~\ref{fig:trace-log} are compiled to
pseudo-assembler if the operation and the guard are compiled separated or if
they are merged.

@@

First a special data
structure called \emph{backend map} is created. This data structure encodes the
mapping from the IR-variables needed by the guard to rebuild the state to the
low-level locations (registers and stack) where the corresponding values will
be stored when the guard is executed.
This data
structure stores the values in a succinct manner using an encoding that uses
8 bits to store 7 bits of information, ignoring leading zeros. This encoding is efficient to create and
provides a compact representation of the needed information in order
to maintain an acceptable memory profile.

Second for each guard a piece of code is generated that acts as a trampoline.
Guards are implemented as a conditional jump to this trampoline in case the
guard check fails.
In the trampoline the pointer to the
backend map is loaded and after storing the current execution state
(registers and stack) execution jumps to a generic bailout handler, also known
as \emph{compensation code},
that is used to leave the compiled trace in case of a guard failure.

Using the encoded location information the bailout handler reads from the
saved execution state the values that the IR-variables had  at the time of the
guard failure and stores them in a location that can be read by the frontend.
After saving the information the control is passed to the frontend signaling
which guard failed so the frontend can read the information passed and restore
the state corresponding to the point in the program.

As in previous sections the underlying idea for the design of guards is to have
a fast on-trace profile and a potentially slow one in the bailout case where
the execution has to return to the interpreter due to a guard failure. At the same
time the data stored in the backend, required to rebuild the state, should be as
compact as possible to reduce the memory overhead produced by the large number
of guards, the numbers in Figure~\ref{fig:backend_data} illustrate that the
compressed encoding currently has about 15\% to 25\% of the size of of the
generated instructions on x86.

@@

The process of compiling a bridge is very similar to compiling a loop.
Instructions and guards are processed in the same way as described above. The
main difference is the setup phase. When compiling a trace we start with a clean
slate. The compilation of a bridge is started from a state (register and stack
bindings) that corresponds to the state during the compilation of the original
guard. To restore the state needed to compile the bridge we use the encoded
representation created for the guard to rebuild the bindings from IR-variables
to stack locations and registers used in the register allocator.  With this
reconstruction all bindings are restored to the state as they were in the
original loop up to the guard. This means that no register/stack reshuffling is
needed before executing a bridge.

@@

The results presented in this section are based on numbers gathered by running
a subset of the standard PyPy benchmarks. The PyPy benchmarks are used to
measure the performance of PyPy and are composed of a series of
micro-benchmarks and larger programs.\footnote{\url{http://speed.pypy.org/}} The
benchmarks were taken from the PyPy benchmarks repository using revision
\texttt{ff7b35837d0f}.\footnote{\url{https://bitbucket.org/pypy/benchmarks/src/ff7b35837d0f}}
The benchmarks were run on a version of PyPy based on the
revision~\texttt{0b77afaafdd0} and patched to collect additional data about the
guards in the machine code
backends.\footnote{\url{https://bitbucket.org/pypy/pypy/src/0b77afaafdd0}} The
tools used to run and evaluate the benchmarks including the patches applied to
the PyPy sourcecode can be found in the repository for this
paper.\footnote{\url{https://bitbucket.org/pypy/extradoc/src/tip/talk/vmil2012}}
All
benchmark data was collected on a MacBook Pro 64 bit running Max OS 10.8 with
the loop unrolling optimization disabled.\footnote{Since loop unrolling
duplicates the body of loops it would no longer be possible to meaningfully
compare the number of operations before and after optimization. Loop unrolling
is most effective for numeric kernels, so the benchmarks presented here are not
affected much by its absence.}

@@

All measurements presented in this section do not take garbage collection of machine code into account. Pieces
of machine code can be globally invalidated or just become cold again. In both
cases the generated machine code and the related data is garbage collected. The
figures show the total amount of operations that are evaluated by the JIT and
the total amount of code and data that is generated from the optimized traces.

@@

\subsection{Space Overhead of Guards}
\label{sub:guard_overhead}
\begin{figure}
    \include{figures/resume_data_table}
    \caption{Resume data sizes}
    \label{fig:resume_data_sizes}
\end{figure}

@@

\begin{figure}
    \include{figures/backend_table}
    \caption{Total size of generated machine code and resume data}
    \label{fig:backend_data}
\end{figure}e.

@@

This low guard failure rate is expected. Most guards do not come from actual
control flow divergences in the user program, but from type checks needed for
type specialization. Various prior work has
shown~\cite{holkner_evaluating_2009, richards_analysis_2010, callau_how_2011}
that most programs in dynamic languages only use a limited amount of runtime
variability. Therefore many guards are needed for making the traces behave
correctly in all cases but fail rarely.

>>>>

\begin{abstract}
Tracing just-in-time (JIT) compilers record linear control flow paths,
inserting operations called guards at points of possible divergence. These
operations occur frequently in generated traces and therefore it is important to
design and implement them carefully to find the right trade-off between
execution speed, deoptimization,
and memory overhead.  In this paper we describe the design decisions about
guards taken in the implementation of the RPython tracing JIT. Furthermore we
measure various properties of guards.
% \o/
\end{abstract}

%___________________________________________________________________________
\todo{mention somewhere that it is to be expected that most guards do not fail}
\todo{better formatting for lstinline}
\section{Introduction}

@@

There are several aspects to be taken into account w.r.t. guards, the first aspect is that due to the
large number of guards the memory overhead related to storing the information
needed for deoptimization should be kept low. A second aspect is that
successfully checking guards, i.e. not leaving the compiled trace, – which is
the common case – should be a cheap operation to execute favouring the on-trace
execution speed in contrast to the deoptimization case where the state has to
be rebuilt using the stored information. These constraints and trade-offs are
what make the design and optimization of guards an important and non-trivial
aspect of the construction of a tracing just-in-time compiler.

%Section~\ref{sec:Evaluation} presents Figures about the absolute number of
%operations for each benchmark, and the overhead produced by the information
%stored at the different levels for the guards
In this paper we want to substantiate the aforementioned observations and
describe based on them the reasoning behind the implementation of guards in
RPython's tracing just-in-time compiler. the contributions of this paper are:
\begin{itemize}
  \item An analysis and benchmark of guards in the context of RPython's tracing JIT.
  %An analysis of guards in the context of RPython's tracing JIT to
  %substantiate the aforementioned observation, based on a set of benchmarks,
  \item detailed measurements about the frequency and the
  overhead associated with guards, and
  \item a description about how guards are implemented in the high\-
  and low-level components of the JIT and describe the rationale behind the design.
\end{itemize}

@@

After profiling identifies an interesting
path, tracing is started thus recording all operations that are executed on this
path. This includes inlining functional calls.
As in most compilers, tracing JITs use an intermediate representation to
store the recorded operations, typically in SSA
form~\cite{cytron_efficiently_1991}. Since tracing follows actual execution the
code that is recorded
represents only one possible path through the control flow graph. Points of
divergence from the recorded path are marked with special operations called
\emph{guards}, these operations ensure that assumptions valid during the
tracing phase are still valid when the code has been compiled and is executed.
In the case of dynamic languages, guards are also used to encode type checks
that come from optimistic type specialization by recording the types of
variables seen during tracing.
After a trace has been recorded it is optimized and then compiled to platform
specific machine code.

When the check of a guard fails, the execution of the machine code must be
stopped and the control is returned to the interpreter, after the interpreter's
state has been restored. If a particular guard fails often a new trace
starting from the guard is recorded. We will refer to this kind of trace as a
\emph{bridge}. Once a bridge has been traced and compiled it is attached to the
corresponding guard by patching the machine code. The next time the guard fails
the bridge will be executed instead of leaving the machine code.

@@

To do this reconstruction it is necessary to take the values of the SSA
variables in the trace to build interpreter stack frames.  Tracing
aggressively inlines functions, therefore the reconstructed state of the
interpreter can consist of several interpreter frames.

If a guard fails often enough, a trace is started from it
to create a bridge, forming a trace tree.
When that happens another use case of resume data
is to reconstruct the tracer state.
After the bridge has been recorded and compiled it is attached to the guard.
If the guard fails later the bridge is executed. Therefore the resume data of
that guard is no longer needed.

There are several forces guiding the design of resume data handling.
Guards are a very common operations in the traces.
However, as will be shown, a large percentage of all operations
are optimized away before code generation.
Since there are a lot of guards
the resume data needs to be stored in a very compact way.
On the other hand, tracing should be as fast as possible,
so the construction of resume data must not take too much time.

@@

Every time a guard is recorded during tracing
the tracer attaches preliminary resume data to it.
The data is preliminary in that it is not particularly compact yet.
The preliminary resume data takes the form of a stack of symbolic frames.
The stack contains only those interpreter frames seen by the tracer.
The frames are symbolic in that the local variables in the frames
do not contain values.
Instead, every local variable contains the SSA variable of the trace
where the value would later come from, or a constant.

@@

After tracing has been finished the trace is optimized.
During optimization a large percentage of operations can be removed.\todo{add a reference to the figure showing the optimization rates?}
In the process the resume data is transformed into its final, compressed form.
The rationale for not compressing the resume data during tracing
is that a lot of guards will be optimized away.
For them, the compression effort would be lost.

@@

Guards interact with optimizations in various ways.
Using many classical compiler optimizations the JIT tries to remove as many
operations, and therefore guards, as possible.
In particular guards can be removed by subexpression elimination.
If the same guard is encountered a second time in a trace,
the second one can be removed.
This also works if a later guard is weaker
and hence implied by an earlier guard.

@@

Storing this additional information is done as follows:
So far, every variable in the symbolic frames
contains a constant or an SSA variable.
After allocation removal the variables in the symbolic frames can also contain
``virtual'' objects.
These are objects that were not allocated so far,
because the optimizer removed their allocation.
The structure of the heap objects that have to be allocated on guard failure
is described by the virtual objects stored in the symbolic frames.
To this end, the content of every field of the virtual object is described
in the same way that the local variables of symbolic frames are described.
The fields of the virtual objects can therefore be SSA variables, constants
or other virtual objects.
They are encoded using the same compact two-byte representation
as local variables.

During the storing of resume data virtual objects are also shared
between subsequent guards as much as possible.
The same observation as about frames applies:
Quite often a virtual object does not change from one guard to the next,
allowing the data structure to be shared.

@@

After the recorded trace has been optimized it is handed over to the platform specific
backend to be compiled to machine code. The compilation phase consists of two
passes over the lists of instructions, a backwards pass to calculate live
ranges of IR-level variables and a forward pass to emit the instructions. During
the forward pass IR-level variables are assigned to registers and stack
locations by the register allocator according to the requirements of the
emitted instructions.  Eviction/spilling is performed based on the live range
information collected in the first pass. Each IR instruction is transformed
into one or more machine level instructions that implement the required
semantics. Operations without side effects whose result is not used are not
emitted. Guards instructions are transformed into fast checks at the machine
code level that verify the corresponding condition.  In cases the value being
checked by the guard is not used anywhere else the guard and the operation
producing the value can merged, further reducing the overhead of the guard.
Figure \ref{fig:trace-compiled} shows how the \lstinline{int_eq} operation
followed by a \lstinline{guard_false} from the trace in Figure~\ref{fig:trace-log} are compiled to
pseudo-assembler if the operation and the guard are compiled separated or if
they are merged.

@@

First a special data
structure called \emph{backend map} is created. This data structure encodes the
mapping from IR-variables needed by the guard to rebuild the state to the
low-level locations (registers and stack) where the corresponding values will
be stored when the guard is executed.
This data
structure stores the values in a succinct manner using an encoding that requires
8 bits to store 7 bits of information, ignoring leading zeros. This encoding is efficient to create and
provides a compact representation of the needed information in order
to maintain an acceptable memory profile.

Second for each guard a piece of code is generated that acts as a trampoline.
Guards are implemented as a conditional jump to this trampoline in case the
guard check fails.
In the trampoline the pointer to the
backend map is loaded and after storing the current execution state
(registers and stack) execution jumps to a generic bailout handler, also known
as \emph{compensation code},
that is used to leave the compiled trace.

Using the encoded location information the bailout handler reads from the
stored execution state the values that the IR-variables had at the time of the
guard failure and stores them in a location that can be read by the frontend.
After saving the information the control is returned to the frontend signaling
which guard failed so the frontend can read the stored information and rebuild
the state corresponding to the point in the program.

As in previous sections the underlying idea for the low-level design of guards is to have
a fast on-trace profile and a potentially slow one in case
the execution has to return to the interpreter. At the same
time the data stored in the backend, required to rebuild the state, should be as
compact as possible to reduce the memory overhead produced by the large number
of guards, the numbers in Figure~\ref{fig:backend_data} illustrate that the
compressed encoding currently has about 15\% to 25\% of the size of of the
generated instructions on x86.

@@

The process of compiling a bridge is very similar to compiling a loop.
Instructions and guards are processed in the same way as described above. The
main difference is the setup phase. When compiling a trace we start with a clean
slate. The compilation of a bridge is started from a state (register and stack
bindings) that corresponds to the state during the compilation of the original
guard. To restore the state needed to compile the bridge we use the backend map
created for the guard to rebuild the bindings from IR-variables
to stack locations and registers.  With this
reconstruction all bindings are restored to the state as they were in the
original loop up to the guard. This means that no register/stack reshuffling is
needed before executing a bridge.

@@

The results presented in this section are based on numbers gathered by running
a subset of the standard PyPy benchmarks. The PyPy benchmarks are used to
measure the performance of PyPy and are composed of a series of
micro-benchmarks and larger programs.\footnote{\url{http://speed.pypy.org/}} The
benchmarks were taken from the PyPy benchmarks repository using revision
\texttt{ff7b35837d0f}.\footnote{\url{https://bitbucket.org/pypy/benchmarks/src/ff7b35837d0f}}
The benchmarks were run on a version of PyPy based on
revision~\texttt{0b77afaafdd0} and patched to collect additional data about
guards in the machine code
backends.\footnote{\url{https://bitbucket.org/pypy/pypy/src/0b77afaafdd0}} The
tools used to run and evaluate the benchmarks including the patches applied to
the PyPy sourcecode can be found in the repository for this
paper.\footnote{\url{https://bitbucket.org/pypy/extradoc/src/tip/talk/vmil2012}}
All
benchmark data was collected on a MacBook Pro 64 bit running Max OS 10.8 with
the loop unrolling optimization disabled.\footnote{Since loop unrolling
duplicates the body of loops it would no longer be possible to meaningfully
compare the number of operations before and after optimization. Loop unrolling
is most effective for numeric kernels, so the benchmarks presented here are not
affected much by its absence.}

@@

All measurements presented in this section do not take garbage collection of resume data and machine code into account. Pieces
of machine code can be globally invalidated or just become cold again. In both
cases the generated machine code and the related data is garbage collected. The
figures show the total amount of operations that are evaluated by the JIT and
the total amount of code and data that is generated from the optimized traces.

@@

\subsection{Overhead of Guards}
\label{sub:guard_overhead}
\begin{figure}
    \include{figures/resume_data_table}
    \caption{Resume data sizes}
    \label{fig:resume_data_sizes}
\end{figure}

@@

\begin{figure}
    \include{figures/backend_table}
    \caption{Total size of generated machine code and resume data}
    \label{fig:backend_data}
\end{figure}

@@
